Simple repression draft

### 2020-05-13 doc/paper.tex
Edits from RP + other small edits

\section{Means}
\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple
repression, let us consider what we can learn from theory and
measurements of fold-change in mean expression. The fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
\label{eq:fc_def}
\end{equation}
where angle brackets denote the averge over a population of cells
and mean mRNA $\langle m\rangle$ is viewed as a function of
repressor copy number $R$.
Also note the implicit assumption in equating ratios of gene
expression, i.e., protein levels, to mRNA levels. The second
equality in~\ref{eq:fc_def} amounts to an assumption that the
translation efficiency, i.e., the number of proteins translated
per mRNA, is the same in both conditions.
This is reasonable since the cells in the two conditions are
identical except for the presence of the transcription factor,
and it is difficult to imagine the transcription factor having
any direct effect on translation.

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
Part of its utility in dissecting transcriptional regulation is
its ratiometric nature, which is always easier than making an
absolute measurement. Also, by measuring otherwise identical cells
with an without a transcription factor present, any biological
noise common to both conditions can be made to cancel away.

Panels C and D of \fig{fig1:means_cartoons} depict a smorgasbord
of mathematized cartoons for simple repression that have appeared
in previous literature. For each cartoon, we calculate the
fold-change in mean gene expression as predicted by that model,
deferring some algebraic details to the appendix.
What we will find is that all cartoons collapse to a single
master curve, shown in \fig{fig1:means_cartoons}B, which contains
just two parameters. We label the parameters $\Delta F_R$, an
effective free energy parametrizing the repressor-DNA
interaction, and $\rho$, which subsumes all details of
transcription in the absence of repressors.
We will offer some intuition for why this master curve exists and discuss
why it is likely impossible to discriminate ``right'' from ``wrong''
cartoons given only measurements of fold-changes in expression.

@@

\subsubsection{Equilibrium, 2-state}
In this simplest model, the promoter is idealized as existing in
one of two states, either repressor bound or repressor unbound.
The rate of transcription, and therefore the level of gene
expression, is assumed to be proportional to the fraction of time
spent in the repressor unbound state.
Mathematically, this amounts to the statement that the mean mRNA level is
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound
state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
copy number, $N_{NS}$ is the number of nonspecific binding sites
in the genome (where repressors spend most of their time when not
bound to the operator, their specifc binding site),
$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
repressor to its operator site. The derivation of this is deferred
to the appendix.

@@

@@

We can greatly simplify the notation, which will be especially
useful for the more complicated models yet to come, by
rexpressing the master equation in vector form. The promoter
states are collected into a vector and the rate constants are
collected into matrices as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:2state_rep_cme}
\end{equation}
where $\matr{I}$ is the identity matrix.
Taking steady state by setting time derivatives to zero,
the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
with the algebra details again deferred to the appendix.
Recall $k_R^+$ is proportional to the repressor copy number, so
in computing fold-change, absence of repressor corresponds to
$k_R^+\rightarrow0$.
Therefore fold-change in this model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.

@@

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
The main subtlety arises since transcription initiation
accompanies a promoter state change, but this matrix can be
understood by analogy to the off-diagonal element of $\matr{K}$.
The master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_{out}} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_{in}} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_{in}} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
                \end{pmatrix},\
\matr{R_{out}} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
                \end{pmatrix}.
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having
$m$ mRNA present along with the promoter having repressor bound,
being empty, or having RNAP bound, respectively.
$\matr{R_{in}}$ describes probability flux into the state
$\vect{p}_m$ from a state with one fewer mRNA, namely
$\vect{p}_{m-1}$, and $\matr{R_{out}}$ describes probability flux
out of the state $\vect{p}_m$ for a state with one more mRNA,
namely $\vect{p}_{m+1}$.
$\matr{K}$ is closely analogous to model 1.

@@

The master equation governing this model is identical to model 2 above,
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_{out}} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_{in}} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\end{equation}
but obviously with different matrices.
The state vector and promoter transition matrix are now
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound,
empty, RNAP closed complex, and RNAP open complex.
Besides increasing dimension by one again, the only new feature in
$\matr{K}$ is the rate $k_O$, representing the (irreversible)
rate of open complex formation from the closed complex.
The initiation matrices are given by
\begin{equation}
\matr{R_{in}} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
                \end{pmatrix},\
\matr{R_{out}} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to model 2.

The expression for mean mRNA is substantially more complicated
now, as worked out in the appendix where we find
\begin{align}
\langle m\rangle &= \frac{r}{\gamma}
        \frac{k_R^- k_P^+ k_O}
        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)}
\\
&= \frac{r}{\gamma}
\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
\end{align}
The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
\begin{align}
\text{fold-change}
&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
\\
&= \left(
        1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
\right)^{-1},
\end{align}
surprisingly reducing to the master curve once again, with
$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
\caption{\textbf{
    An overview of the simple repression motif at the level of means.}
    (A) depicts the qualitative biological picture of simple repression.
    (B) shows the ``master curve'' to which all cartoons in (C) and (D) collapse.
    (C) shows two possible equilibrium pictures of simple repression, while
    (D) shows five possible coarse-grainings in a nonequilibrium picture.
\label{fig1:means_cartoons}
\end{figure}

>>>>

\section{Means}
\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple
repression, let us consider what we can learn from theory and
measurements of fold-change in mean expression. The fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
\label{eq:fc_def}
\end{equation}
where angle brackets denote the averge over a population of cells
and mean mRNA $\langle m\rangle$ is viewed as a function of
repressor copy number $R$.
The second
equality in~\ref{eq:fc_def} follows from assuming that the
translation efficiency, i.e., the number of proteins translated
per mRNA, is the same in both conditions.
In other words, we assume that mean protein level is proportional
to mean mRNA level, and that the proportionality constant is the
same in both conditions and therefore cancels out in the ratio.
This is reasonable since the cells in the two conditions are
identical except for the presence of the transcription factor,
and it is difficult to imagine the transcription factor having
any direct effect on translation.

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
Part of its utility in dissecting transcriptional regulation is
its ratiometric nature, which is always easier than making an
absolute measurement. Also, by measuring otherwise identical cells
with and without a transcription factor present, any biological
noise common to both conditions can be made to cancel away.

\fig{fig1:means_cartoons} depicts a smorgasbord
of mathematized cartoons for simple repression that have appeared
in previous literature. For each cartoon, we calculate the
fold-change in mean gene expression as predicted by that model,
deferring some algebraic details to the appendix.
What we will find is that all cartoons collapse to a single
master curve, shown in \fig{fig1:means_cartoons}c, which contains
just two parameters. We label the parameters $\Delta F_R$, an
effective free energy parametrizing the repressor-DNA
interaction, and $\rho$, which subsumes all details of
transcription in the absence of repressors.
We will offer some intuition for why this master curve exists and discuss
why it is likely impossible to discriminate ``right'' from ``wrong''
cartoons given only measurements of fold-changes in expression.

@@

\subsubsection{Equilibrium, 2-state}
In this simplest model, the promoter is idealized as existing in
one of two states, either repressor bound or repressor unbound.
The rate of transcription, and therefore the level of gene
expression, is assumed to be proportional to the fraction of time
spent in the repressor unbound state.
Mathematically, this amounts to the statement that the mean mRNA level is
given by
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound
state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
copy number, $N_{NS}$ is the number of nonspecific binding sites
in the genome (where repressors spend most of their time when not
bound to the operator, their specifc binding site),
$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
repressor to its operator site. The derivation of this result is deferred
to the appendix.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
\caption{
\textbf{
An overview of the simple repression motif at the level of means.}
(A) depicts the qualitative biological picture of simple repression.
(B) shows a variety of possible mathematized cartoons of simple
repression, along with the corresponding complexity $\rho$ of the
architecture, an effective parameter which subsumes all details
of the architecture excluding the repressor.
(C) shows the ``master curve'' to which all cartoons in (B) collapse.
}
\label{fig1:means_cartoons}
\end{figure}

@@

We can greatly simplify the notation, which will be especially
useful for the more complicated models yet to come, by
re-expressing the master equation in vector form. The promoter
states are collected into a vector and the rate constants are
collected into matrices as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R} \vect{p}_{m-1}(t)
                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
                - \gamma m \matr{I} \vect{p}_m(t),
\label{eq:2state_rep_cme}
\end{equation}
where $\matr{I}$ is the identity matrix.
Taking steady state by setting time derivatives to zero,
the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
with the algebra details again deferred to the appendix.
Recall $k_R^+$ is proportional to the repressor copy number, so
in computing fold-change, absence of repressor corresponds to
$k_R^+\rightarrow0$.
Therefore fold-change in this model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.

@@

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
The main subtlety arises since transcription initiation
accompanies a promoter state change.
This can be understood by analogy to $\matr{K}$. The
off-diagonals and diagonals of $\matr{K}$ correspond to
transitions arriving at or departing from, respectively, the
promoter state of interest. If transcription initiations are
accompanied by promoter state changes, we must have separate
matrices for arriving and departing transcription events since
the arriving and departing transitions have different initial
copy number of mRNA, unlike for $\matr{K}$ where they are the same.
The master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
                - \gamma m \matr{I} \vect{p}_m(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having
$m$ mRNA present along with the promoter having repressor bound,
being empty, or having RNAP bound, respectively.
$\matr{R_A}$ describes probability flux \textit{arriving} at the state
$\vect{p}_m$ from a state with one fewer mRNA, namely
$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
\textit{departing} from the state $\vect{p}_m$
for a state with one more mRNA, namely $\vect{p}_{m+1}$.
$\matr{K}$ is closely analogous to model 1.

@@

The master equation governing this model is
identical in form to model 2 above, namely
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t)
                - \gamma m \matr{I} \vect{p}_m(t),
\end{equation}
but obviously with a higher-dimensional state space and different matrices.
The state vector and promoter transition matrix are now
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound,
empty, RNAP closed complex, and RNAP open complex.
Besides increasing dimension by one again, the only new feature in
$\matr{K}$ is the rate $k_O$, representing the (irreversible)
rate of open complex formation from the closed complex.
The initiation matrices are given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to model 2.

The expression for mean mRNA is substantially more complicated
now, as worked out in the appendix where we find
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+ k_O}
        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
\end{equation}
which can be simplified to
\begin{equation}
\langle m\rangle
= \frac{r}{\gamma}
\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
\end{equation}
The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
\begin{align}
\text{fold-change}
&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
\\
&= \left(
        1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
\right)^{-1},
\end{align}
surprisingly reducing to the master curve once again, with
$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.

@@
